Reduced diversity and increased specialzation of Alaska salmon fishermen:

### 2016-10-05 salmon/paper/salmon_paper.Rmd
salmon update

---
title: "Reduced diversity and increased specialzation of Alaska salmon fishermen: effects on variation in individual revenues"
author: ''
date: ''
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document: default
---

@@

Previous analyses have largely focused on patterns of diversification across ecosystems and the benefits to adopt more or less diverse fishing strategies (Kasperski and Holland 2013; Anderson et al. in review). Fisheries in Alaska have been studied extensively, because they are one of the most productive regions in North America (both by landed weight, and revenue cite) and they allow individuals to adopt a wide range of fishing strategies. Anderson et al. (in review) defined strategies as a permit of collection of permits held by single individuals. While adding new permits allows fishers to increase species diveristy, few individual permits allow individuals to harvest more than one species (examples include halibut, sablefish, herring, crab, shellfish, shrimp, sea cucumber). Exceptions to these single species permits incldue participants in federally managed groundfish fisheries (CFEC 'M' permits) and participants in state managed salmon fisheries (CFEC 'S' permits).

Understanding the effects of species diversification on salmon fishers is particularly important, as the number of salmon specialists (defined as fishers holding a single salmon permit) has increased over the last 30 years as a fraction of the total (Anderson et al. in review). For these limited entry salmon fisheries, general trends that have emerged are that more people are participating since the early 2000s (Fig. A1), median revenue is increasing (across individuals, Fig. A2), and landed species diversity declining over time (averaged across indivduals, Fig. A6). Across different salmon fisheries, other than a slight negative effect of diversification on variability (similar to Anderson et al. in review), there appears to be little correlation between diversity and revenue (Fig. A4). In terrestrial agriculture, there are numerous examples where diversification may affect the mean or CV but not both (Purdy, Barry M; Langemeier, Michael R; Featherstone, Allen M. Journal of Agricultural and Applied Economics29.1 (Jul 1, 1997): 149.), and the same result may be true for salmon fishers. Further, these impacts of diversification may be variable over time.

Five species of salmon are encountered in Alaska waters (Chinook, chum, pink, sockeye, coho), and the ability of individual salmon fishers to diversify across species is in part a function of geography and gears. For example, salmon may be caught with purse seine gear (n = 6 areas), drift gillnets (5 areas), and set gillnets (12 areas). Other gears include beach seine and troll gears (though these are generally restricted to Kodiak and Southeast Alaska, respectively). While some of these combinations of gears and areas allow fishers to diversify, there are many examples of permits where the catch is dominated by a single species. Perhaps the best examples include fisheries that target sockeye salmon (e.g. Bristol Bay set and drift gillnet fisheries). For these fisheries, specialization is associated with higher revenue and reduced CV in annual revenue (Anderson et al. in review).

A secondary motivation for focusing on the effects of diversification for salmon fishers is to better understand the long term consequences of hatchery salmon production on the revenue and variability in revenue of individuals. Following low returns and declining productivity in the mid 1980s, many hatchery programs in the state rapidly increased hatchery releases (Mahnken et al. 1998). Many of these hatchery programs increased production to stabilize the variability of fishers catch (and revenue), though these programs may also have costs to wild production (Ruggerone et al. 2015). [Other history here?]

In addition to the changing species composition of fish avaialble to be harvested, the benefits (or costs) of diversification are likely affected by a number of exogenous drivers, including marketing, prices, new processing technology, etc (cite Gunnar Knapp + others). [Break out as own paragraph+add more examples?] Following price crashes in the early 2000s, many of these factors have been positive for salmon fishermen (increasing trends in prices and catches, Fig. A3, A12). In contrast, the late 1980s - early 1990s experienced high abundance and low prices (in part due to collapses of foreign currencies). Combined with the increased trend in hatchery production and resulting change species compositon of adult salmon, these external factors may have also altered incentives for diversification. If there's an incentive to substitute high value species (Chinook, sockeye) with abundant low value species (pink, chum), the analog in the terrrestrial framring world would be crop substitution of low-volume high value crops (spices) with high-volume low value crops (cereals) (other examples in literature = cocoa, tobacco).

The objective of our analysis is to better understand the effects of diversification on single-permit salmon holders by extending the methods used by Anderson et al. (in review) to include time-varying effects. These time-varying effects will allow us to compare temporal trends across regions, identify periods when having a diverse catch portfolio was most beneficial, and allow us to evaluate support for effects of longer term drivers in the region.

@@

We obtained fisheries landings and revenue data for all permit holders in Alaska from 1985 - 2014 from the Alaskan Commercial Fisheries Entry Commission (CFEC). We adjusted for inflation by setting all revenues in 2009 USD. Because were were only interested in single permit fisheries, we adopted a simplified version of the filtering steps described in  Anderson et al. (in review). After aggregating individual fish tickets to annual data, we retained individuals who only fished one of the six salmon permits included in a year (Table 1, Fig. A5). Like Anderson et al. (in review) we also worked with a log-differenced dataset to remove non-stationary individuals; to avoid complications of individuals switching strategies, we only included pairs of years for a given person where they did not change strategies.

@@

Hatchery release trends for some regions included in our data have been summarized previously (e.g. Ward et al. in review summarized the trends for Prince William Sound). Release patterns have been dominated by pink and chum salmon (for PWS, these species represent 80% and 18% of total releases for 2012-2014 respectively, Ward et al. in review) and production has increased since the late 1970s (217 million juvenile pink salmon in PWS in 1985 versus 672 million in 2014). Though our analysis doesn't include , as a contrast we can do coarse comparisons across regions and species to examine potential effects of hatchery fish on variability. For example, we can compare the variation in pink salmon returns (earnings or weight) from Prince William Sound (largely hatchery) to those from Southeast Alaska (largely wild). Similarly, we can compare the variability in earnings from species dominated by hatchery returns (pink, chum) to those with less hatchery influence (Chinook, sockeye). To make these comparisons, we aggregated the earnings and landings by species and permit, then calculated 3-year rolling CVs. As a second comparison, we examined whether long-term participants in any of the fisheries had experienced a change in variability. Similar to Kasperski and Holland (2013) who calculated CVs of vessels before and after significant managment events, we identified individuals who fished the same permit for at least three years in the first decade of our data (1985-1995) and at least three years in the last decade of the data (2004-2014).

@@

Anderson et al. (in review) implemented a hierarchical Bayesian model that allowed group level predictors and covariates (diversity of catch, days fished) to simultaneously affect the mean and the variance of revenue for individual fishers. This model also included random intercepts by strategy and year; these terms allowed all permit holders in a given year to experience common fluctuations (external drivers including price changes, fluctuations in fish population abundance, or environmental disasters such as EVOS).

To ask whether how benefits of diversification may have changed over time, we extend the model of Anderson et al. (in review) to also include time-varying random effects in the slope coefficients of diversification. We define the variables $\Delta { S }_{ t }=log\left( { S }_{ t }/{ S }_{ t-1 } \right)$ to represent the percent change in species diversity of the catch between years and $\Delta { D }_{ t }=log\left( { D }_{ t }/{ D }_{ t-1 } \right)$ to represent the percent change in fishing effort (days fished) between years. The model for the the expected log revenue of individual *i* in year *t*, $log({ R }_{ i,t })$, can be written as $$\widehat { log\left( { R }_{ i,t } \right)  } ={ B }_{ 0,j,t }+{ B }_{ 1,j,t }\cdot \Delta S_{ t }+{ B }_{ 2,j,t }\cdot \left( \Delta S_{ t }\cdot { S }_{ t-1 } \right) +{ B }_{ 3,j,t }^{ 2 }\cdot \left( \Delta S_{ t }\cdot { S }_{ t-1 }^{ 2 } \right) +{ B }_{ 4,j }\cdot \Delta D_{ t }+log\left( { R }_{ i,t-1 } \right) $$, where *j* subscripts the salmon permit (Table 1). We model the time varying permit specific intercepts as random and exchangeable across years, ${ B }_{ 0,j,t }\sim Normal\left( { u }_{ { B }_{ 0 } },{ \sigma  }_{ { B }_{ 0 } } \right)$. The term ${ B }_{ 1,j,t }$ represents the permit level random effect of diversifying species. Our model extendeds that of Anderson et al. (in review) to include the interaction terms $\Delta S_{ t }\cdot { S }_{ t-1 }$ which allow the effect of diversification to be dependent on the previous level of diversification. The squared term allows this interaction to be quadratic, and the coefficients ${ B }_{ 2,j,t }$ and ${ B }_{ 3,j,t }$, allow this to vary by permit and through time. The flexibility in these terms allows the benefit of increasing species diversity to be positive in some years (generalizing benefits revenue), negative in other years (specializing benefits revenue) and neutral in other years (revenue is independent alterations to catch composition). We constrain the random effects associated with diversification to be modeled as autoregressive processes with shared variances across strategies, for example: ${ B }_{ 1,j,t }\sim Normal\left( { B }_{ 1,j,t-1 },{ \sigma  }_{ { B }_{ 1 } } \right)$. Like Anderson et al. (in review) we also allow changing effort (measured as percent change in days fished) ${ B }_{ 4,j }$ to vary by strategy, but we do not allow this effect to vary through time. Finally, the term $log\left( { R }_{ i,t-1 } \right)$ is included as an offset (Anderson et al. in review).

- Fit models from Sean's paper, but include time varying interactions on species diversity.  GLMM code below

@@

@@

- Prices are generally synchronous for all spp (especially coho/chum/pink), and have increased at a faster rate compared to chum and pink salmons since 2000 (Fig A8-A9). We're also seeing a trend toward greater synchrony between species (Fig A11-A12, A14). Projecting future prices are complex -- Gunnar Knapp has a good presentation on this http://www.adfg.alaska.gov/static/fishing/PDFs/commercial/gk_trends_4-12-12.pdf. Synchrony in prices may largely be function of marketing, improved product, freezing, etc.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
revenueThreshold = 10000

dat.annual = readRDS(file="../../portfolio/data-generated/cfec-annual-for-modeling.rds")

# single permit holders, salmon permits
dat.annual  = filter(dat.annual ,npermit==1 & substr(strategy_permit,1,1)=="S") %>%
  group_by(strategy_permit) %>% 
  mutate(nPeople = length(unique(p_holder))) %>% 
  filter(nPeople > 200) %>% 
  dplyr::select(-nPeople)
# drop troll
dat.annual = dat.annual[dat.annual$strategy_permit%in%c("S05B","S15B")==FALSE,]
# drop S04X, Y, Z because low participation (< 20) in many years
# S04W deleted because low median earnings
dat.annual = dat.annual[dat.annual$strategy_permit%in%c("S04W", "S04X","S04Y","S04Z")==FALSE,]

agg.annual = group_by(dat.annual, strategy_permit, year) %>% 
  summarize(people = length(unique(p_holder)),
    specDiv = mean(specDiv),
    meanrev = median(revenue),
    low = quantile(revenue,0.25),
    upp = quantile(revenue,0.75))

dat.diff = readRDS(file="../../portfolio/data-generated/cfec-diff-for-modeling.rds")
dat.diff = dat.diff[which(dat.diff$revenue >= revenueThreshold & dat.diff$revenue.prev >= revenueThreshold), ]
dat.diff = dat.diff[dat.diff$strategy_permit%in%c("S01A","S01E","S01K","S01L","S01M",
  "S03A","S03E","S03H","S03M","S03T","S04D","S04H","S04K","S04M","S04T"),]

agg.diff = group_by(dat.diff, strategy_permit) %>% 
  summarize(specDiv = mean(specDiv), meanrev = mean(revenue),
    cv = sd(log(revenue/revenue.prev)))

# Also load in raw CFEC data for plots
cfec <- feather::read_feather("../../portfolio/data-generated/cfec.feather")
cfec = cfec[cfec$p_fshy%in%c("S 01A", "S 01E", "S 01K", "S 03A", "S 03E", "S 01M"),]
cfec$year = as.numeric(substr(cfec$landdate, 1, 4))

# only use people who are using just that permit - diversification within permit
cfec.raw = cfec
dat.salmon = dat.diff[dat.diff$strategy_permit%in%c("S01A", "S01E", "S01K", "S03A", "S03E","S01M"),]

# Restrict analysis to people who keep same strategy - complicates days, etc
dat.salmon = dat.salmon[which(dat.salmon$strategy_permit==dat.salmon$strategy_permit.prev),]

dat.salmon$days.change = log((dat.salmon$days_permit+1) / (dat.salmon$days_permit.prev+1))
dat.salmon$year_permit = as.factor(paste(dat.salmon$year, dat.salmon$strategy_permit))

cfec = cfec[which(paste(cfec$p_holder,cfec$year) %in% paste(dat.salmon$p_holder,dat.salmon$year)),]

# Make come plots for the appendix of (1) median fishing dates over time, (2) length of fishing season over time (calculated as width of 95% intervals)
cfec.raw$month = as.numeric(substr(cfec.raw$day, 1,2))
cfec.raw$day = as.numeric(substr(cfec.raw$day, 4,5))
cfec.raw$jday = mdy.date(cfec.raw$month, cfec.raw$day, cfec.raw$year) - mdy.date(1, 1, cfec.raw$year)
# summarize the temporal distribution of the catches
seasons = group_by(cfec.raw, p_fshy, year) %>% 
  summarize(q = mean(sample(jday, size = 10000, replace=T, prob=g_earn)),
    lower = quantile(sample(jday, size = 10000, replace=T, prob=g_earn),0.025),
    upper = quantile(sample(jday, size = 10000, replace=T, prob=g_earn),0.975),
    length = upper - lower)
seasons = rename(seasons, permit = p_fshy)
appendixPlot0_p1 = ggplot(seasons, aes(year, q, color = permit, group=permit)) + geom_line() + xlab("Year") + ylab("Mean julian day")
appendixPlot0_p2 = ggplot(seasons, aes(year, length, color = permit, group=permit)) + geom_line() + xlab("Year") + ylab("Season length")
```

```{r, echo = FALSE, eval=FALSE}
dat = dat.salmon
# create other category, lumping coho/chinook
dat$salm = dat$coho+dat$chnk
dat$salm.prev = dat$coho.prev+dat$chnk.prev

dat$totsalm = dat$salm + dat$chum + dat$sock + dat$pink
dat$totsalm.prev = dat$salm.prev + dat$chum.prev + dat$sock.prev + dat$pink.prev

dat$salm = dat$salm/dat$totsalm
dat$sock = dat$sock/dat$totsalm
dat$pink = dat$pink/dat$totsalm
dat$chum = dat$chum/dat$totsalm
dat$chnk = dat$chnk/dat$totsalm
dat$coho = dat$coho/dat$totsalm
dat$salm.prev = dat$salm.prev/dat$totsalm.prev
dat$sock.prev = dat$sock.prev/dat$totsalm.prev
dat$pink.prev = dat$pink.prev/dat$totsalm.prev
dat$chum.prev = dat$chum.prev/dat$totsalm.prev
dat$chnk.prev = dat$chnk.prev/dat$totsalm.prev
dat$coho.prev = dat$coho.prev/dat$totsalm.prev

dat = dat[-which(is.na(dat$salm+dat$sock+dat$pink+dat$chum)),]
dat$logdiff = log(dat$revenue/dat$revenue.prev)

dat_old = dat
dat$sumP2 = abs(dat$salm - dat$salm.prev) + 
  abs(dat$pink-dat$pink.prev) + 
  abs(dat$chum-dat$chum.prev) + 
  abs(dat$sock-dat$sock.prev)
dat = dat[which(dat$sumP2 < 0.3),]
```

```{r, echo = FALSE, eval=FALSE, fig.cap="Results for PWS Purse seine fishery", fig.pos="placeHere", cache=TRUE}

df = dat[which(dat$strategy_permit=="S01E"),]
pamk_s01e <- pamk(df[,c("chum","pink","sock","salm")])
df$group = c("Pink-chum","Pink")[pamk_s01e$pamobject$clustering]

p3 = ggtern(data=df[,c("chum","pink","sock","group")], aes(x=chum,y=pink,sock,color=as.factor(group))) +
  geom_point(alpha=0.3) +
  labs(x="CHUM",y="PINK",z="SOCK") + theme(
        axis.title=element_text(size=8)) + theme(legend.position = "none")

p0 = group_by(df, year, group) %>% 
  summarize(n = length(unique(p_holder))) %>% 
ggplot(aes(x = year, y = n, group=group, col = group)) + geom_line() + ylab("People") + ggtitle("S01E")

p1 = group_by(df, year, group) %>% 
  summarize(meanrev = log(mean(revenue + revenue.prev)/2), 
    cv = sd(logdiff,na.rm=T)) %>% 
ggplot(aes(x = year, y = meanrev, group=group, col = group)) + geom_line() + ylab("Mean revenue") + ggtitle("S01E")

p2 = group_by(df, year, group) %>% 
  summarize(meanrev = log(mean(revenue + revenue.prev)/2), 
    cv = sd(logdiff,na.rm=T)) %>% 
ggplot(aes(x = year, y = cv, group=group, col = group)) + geom_line() + ylab("CV revenue") + ggtitle("S01E")

g = group_by(df, year, group) %>% 
  summarize(meanrev = log(mean(revenue + revenue.prev)/2), 
    cv = sd(logdiff,na.rm=T))

gridExtra::grid.arrange(p3,p0,p1,p2,ncol=2)
```

```{r makePlots, echo = FALSE}
appendixPlot1 = ggplot(agg.annual, aes(year, people, colour = specDiv)) + geom_line() + facet_wrap(~ strategy_permit, scale="free") + ylab("Participation") + xlab("Year")

appendixPlot2_p1 = ggplot(agg.diff, aes(specDiv, log(meanrev), colour = cv, label=strategy_permit)) + xlab("Species diversity") +ylab("Ln (mean revenue)") + geom_text(size=2)
appendixPlot2_p2 = ggplot(agg.diff, aes(specDiv, cv, colour = log(meanrev), label=strategy_permit)) + xlab("Species diversity") +ylab("CV") + geom_text(size=2)
appendixPlot2_p3 = ggplot(agg.diff, aes(log(meanrev), cv, colour = specDiv, label=strategy_permit)) + ylab("CV") + xlab("Ln (mean revenue)") + geom_text(size=2)
```

```{r, fig.cap="Species composition for 1988 and 2014 (each dot is a person) in the S03E fishery", fig.pos="placeHere", echo=FALSE, fig.height = 2, eval=FALSE}
### One way to visualize the change is with ternary plots over the 3 axes (chum / sockeye / pink)

daty = group_by(cfec[cfec$p_fshy=="S 03E" & cfec$spec %in% c("CHUM","PINK","SOCK","COHO","CHNK"),], spec, p_holder, year) %>%
  summarize(p = sum(g_earn)) %>%
  group_by(p_holder, year) %>% mutate(p = p/sum(p))
daty <- melt(daty, id.vars = c("spec", "p_holder", "year"))
daty <- dcast(daty, p_holder + year ~ spec + variable)[,-1]
daty[is.na(daty)]=0
ggtern(data=daty[which(daty$year%in%c(1988,2014)),c("year", "CHUM_p","PINK_p","SOCK_p")], aes(x=CHUM_p,y=PINK_p,SOCK_p,color=as.factor(year))) +
  geom_point(alpha=0.3,size=1) +
  labs(x="CHUM",y="PINK",z="SOCK",title="Proportion of annual revenue (S03E)")
```

```{r, echo=FALSE, eval=FALSE}

mu.mod = glmmTMB(log(revenue) ~ specDiv.change + specDiv.change:specdiv.prev + specDiv.change:I(specdiv.prev^2) + days.change + (1+days.change|strategy_permit) + (-1+specDiv.change + specDiv.change:specdiv.prev + specDiv.change:I(specdiv.prev^2)|year_permit), offset = log(revenue.prev), dat.salmon)

cv.mod = glmmTMB(log(abs(residuals(mu.mod))) ~ specDiv.change + specDiv.change:specdiv.prev + specDiv.change:I(specdiv.prev^2) + days.change + (1+days.change|strategy_permit) + (-1+specDiv.change + specDiv.change:specdiv.prev + specDiv.change:I(specdiv.prev^2)|year_permit), dat.salmon)
# to ask whether benefit of diversificaiton is changing via strategy, we want to summarize (1) direction of 
# benefit (e.g. is the effect negative or positive), and (2) the magnitude (how steep / shallow it is). With the
# quadratic model, 
# 'squared term' indicates concave up/down
# derivative is magnitude, 2a * x + b

df = ranef(mu.mod)$cond$year_permit
colnames(df) = c("c","b","a")
df$year = as.numeric(substr(rownames(df), 1, 4))
df$strategy = substr(rownames(df), 6, 9)
df$deriv = 2*df$a + df$b
# also calculate change of going from 1 to 2 spp
df$pred1 = df$a*(1^2) + df$b*1 + df$c + fixef(mu.mod)$cond["specDiv.change"] + fixef(mu.mod)$cond["specDiv.change:specdiv.prev"]*1
# add fixed effects into predictions

# We can demonstrate the quadratic curves by year for a single strategy / set of years
df.s01e = df[df$strategy=="S01E",]
demo = expand.grid("speciesDiversity" = seq(1,4,length.out=100), "year" = 1986:2014)
demo = left_join(demo, df.s01e)
demo$effect = demo$a * (demo$speciesDiversity^2) + demo$b * (demo$speciesDiversity) + demo$c
```

```{r, echo = FALSE, warning=FALSE, message=FALSE, results='hide', eval=FALSE}
pdf("../../salmon/Salmon_residuals.pdf")
# Residuals for mean model
dat.salmon$mu.resid = residuals(mu.mod)
dat.salmon$mu.fitted = fitted(mu.mod)

ggplot(dat.salmon, aes(mu.fitted, mu.resid, group = strategy_permit)) + geom_point(alpha=0.05,size=1) + facet_wrap(~strategy_permit) + xlab("Fitted") + ylab("Residuals") + ggtitle("Mean model")
ggplot(dat.salmon, aes(specDiv.change, mu.resid, group = strategy_permit)) + geom_point(alpha=0.05,size=1) + facet_wrap(~strategy_permit) + xlab("Change in species diversification") + ylab("Residuals") + ggtitle("Mean model") + geom_hline(aes(yintercept=0),color="salmon1")
ggplot(dat.salmon, aes(specdiv.prev, mu.resid, group = strategy_permit)) + geom_point(alpha=0.05,size=1) + facet_wrap(~strategy_permit) + xlab("Previous species diversity") + ylab("Residuals") + ggtitle("Mean model") + geom_hline(aes(yintercept=0),color="salmon1")
ggplot(dat.salmon, aes(log((revenue+revenue.prev)/2), mu.resid, group = strategy_permit)) + geom_point(alpha=0.05,size=1) + facet_wrap(~strategy_permit) + xlab("Average revenue") + ylab("Residuals") + ggtitle("Mean model") + geom_hline(aes(yintercept=0),color="salmon1")
ggplot(dat.salmon, aes(specDiv.change, log(abs(mu.resid)), group = strategy_permit)) + geom_point(alpha=0.05,size=1) + facet_wrap(~strategy_permit) + xlab("Change in species diversification") + ylab("Ln abs(residuals)") + ggtitle("Mean model")
ggplot(dat.salmon, aes(specdiv.prev, log(abs(mu.resid)), group = strategy_permit)) + geom_point(alpha=0.05,size=1) + facet_wrap(~strategy_permit) + xlab("Previous species diversity") + ylab("Ln abs(residuals)") + ggtitle("Mean model")
ggplot(dat.salmon, aes(specdiv.prev, log(abs(mu.resid)), group = strategy_permit)) + geom_point(alpha=0.05,size=1) + facet_wrap(~strategy_permit) + xlab("Previous species diversity") + ylab("Ln abs(residuals)") + ggtitle("Mean model")

# Residuals for CV model
dat.salmon$cv.resid = residuals(cv.mod)
dat.salmon$cv.fitted = fitted(cv.mod)

>>>>

---
title: 'Reduced diversity and increased specialzation of Alaska salmon fishermen:
  effects on variation in individual revenues'
author: ''
date: ''
output:
  word_document: default
  pdf_document:
    fig_caption: yes
  html_document: default
bibliography: refs.bib
---

@@

Previous analyses have largely focused on patterns of diversification across ecosystems and the benefits to adopt more or less diverse fishing strategies [@kasperski_income_2013; @anderson_benefits]. Fisheries in Alaska have been studied extensively, because they are one of the most productive regions in North America (both by landed weight, and revenue [@office_of_science_and_technology_fisheries_2016]) and they allow individuals to adopt a wide range of fishing strategies. [@anderson_benefits] defined strategies as a permit of collection of permits held by single individuals. While adding new permits allows fishers to increase species diveristy, few individual permits allow individuals to harvest more than one species (examples of single-species permits in the Gulf of Alaska include halibut, sablefish, herring, crab, shellfish, shrimp, sea cucumber). Some exceptions to these single species permits include participants in federally managed groundfish fisheries and participants in state managed salmon fisheries.

Understanding the effects of species diversification on salmon fishers is particularly important, as the number of salmon specialists (defined as fishers holding a single salmon permit) has increased over the last 30 years as a fraction of the total [@anderson_benefits]. For these limited entry salmon fisheries, general trends that have emerged are that more people are participating since the early 2000s (Fig. A1), median revenue is increasing (across individuals, Fig. A2), and landed species diversity declining over time (averaged across indivduals, Fig. A6). Across different salmon fisheries, other than a slight negative effect of diversification on variability (similar to [@anderson_benefits]), there appears to be little correlation between diversity and revenue (Fig. A4). In terrestrial agriculture, there are numerous examples where diversification may affect the mean or CV but not both [@purdy_financial_1997], and the same result may be true for salmon fishers. Further, these impacts of diversification may be variable over time.

Five species of salmon are encountered in Alaska waters (Chinook, chum, pink, sockeye, coho), and the ability of individual salmon fishers to diversify across species is in part a function of geography and gears. For example, salmon may be caught with purse seine gear (n = 6 areas), drift gillnets (5 areas), and set gillnets (12 areas). Other gears include beach seine and troll gears (though these are generally restricted to Kodiak and Southeast Alaska, respectively). While some of these combinations of gears and areas allow fishers to diversify, there are many examples of permits where the catch is dominated by a single species. Perhaps the best examples include fisheries that target sockeye salmon (e.g. Bristol Bay set and drift gillnet fisheries). For these fisheries, specialization is associated with higher revenue and reduced CV in annual revenue [@anderson_benefits].

A secondary motivation for focusing on the effects of diversification for salmon fishers is to better understand the long term consequences of hatchery salmon production on the revenue and variability in revenue of individuals. Following low returns and declining productivity in the mid 1980s, many hatchery programs in the state rapidly increased hatchery releases [@mahnken_historical_1998; @leber_stock_2008]. Many of these hatchery programs increased production to stabilize the variability of fishers catch (and revenue), though these programs may also have costs to wild production [@ruggerone_productivity_2015]. The intra- and inter-specific effects of large scale hatchery programs has been debated [@hilborn_review_2001; @hilborn_hatcheries_1992; @wertheimer_comment:_2001; @heard_hatchery], but the effects of hatcheries have largely been focused on ecological interactions of aggregate yield to the fishery [@leber_stock_2008] or economic impacts [boyce_alaska_1993], rather than to variability experienced by individual fishermen. [Other history here?]

In addition to the changing species composition of fish avaialble to be harvested, the benefits (or costs) of diversification are likely affected by a number of exogenous drivers, including marketing, prices, new processing technology [@knapp_trends]. [Break out as own paragraph+add more examples?] Following price crashes in the early 2000s, many of these factors have been positive for salmon fishermen (increasing trends in prices and catches, Fig. A3, A12). In contrast, the late 1980s - early 1990s experienced high abundance and low prices (in part due to collapses of foreign currencies). Combined with the increased trend in hatchery production and resulting change species compositon of adult salmon, these external factors may have also altered incentives for diversification. If there's an incentive to substitute high value species (Chinook, sockeye) with abundant low value species (pink, chum), the analog in the terrestrial framring world would be crop substitution of low-volume high value crops (spices) with high-volume low value crops (cereals) (other examples in literature = cocoa, tobacco).

The objective of our analysis is to better understand the effects of diversification on single-permit salmon holders by extending the methods used by [@anderson_benefits] to include time-varying effects. These time-varying effects will allow us to compare temporal trends across regions, identify periods when having a diverse catch portfolio was most beneficial, and allow us to evaluate support for effects of longer term drivers in the region.

@@

We obtained fisheries landings and revenue data for all permit holders in Alaska from 1985 - 2014 from the Alaskan Commercial Fisheries Entry Commission (CFEC). We adjusted for inflation by setting all revenues in 2009 USD. Because were were only interested in single permit fisheries, we adopted a simplified version of the filtering steps described in [@anderson_benefits]. After aggregating individual fish tickets to annual data, we retained individuals who only fished one of the six salmon permits included in a year (Table 1, Fig. A5). Like [@anderson_benefits] we also worked with a log-differenced dataset to remove non-stationary individuals; to avoid complications of individuals switching strategies, we only included pairs of years for a given person where they did not change strategies.

@@

Hatchery release trends for some regions included in our data have been summarized previously (e.g. [@ward_evaluating] summarized the trends for Prince William Sound). Release patterns have been dominated by pink and chum salmon (for PWS, these species represent 80% and 18% of total releases for 2012-2014 respectively, [@ward_evaluating]) and production has increased since the late 1970s (217 million juvenile pink salmon in PWS in 1985 versus 672 million in 2014). Though our analysis doesn't include , as a contrast we can do coarse comparisons across regions and species to examine potential effects of hatchery fish on variability. For example, we can compare the variation in pink salmon returns (earnings or weight) from Prince William Sound (largely hatchery) to those from Southeast Alaska (largely wild). Similarly, we can compare the variability in earnings from species dominated by hatchery returns (pink, chum) to those with less hatchery influence (Chinook, sockeye). To make these comparisons, we aggregated the earnings and landings by species and permit, then calculated 3-year rolling CVs. As a second comparison, we examined whether long-term participants in any of the fisheries had experienced a change in variability. Similar to [@kasperski_income_2013] who calculated CVs of vessels before and after significant managment events, we identified individuals who fished the same permit for at least three years in the first decade of our data (1985-1995) and at least three years in the last decade of the data (2004-2014).

@@

[@anderson_benefits] implemented a hierarchical Bayesian model that allowed group level predictors and covariates (diversity of catch, days fished) to simultaneously affect the mean and the variance of revenue for individual fishers. This model also included random intercepts by strategy and year; these terms allowed all permit holders in a given year to experience common fluctuations (external drivers including price changes, fluctuations in fish population abundance, or environmental disasters such as EVOS).

To ask whether how benefits of diversification may have changed over time, we extend the model of [@anderson_benefits] to also include time-varying random effects in the slope coefficients of diversification. We define the variables $\Delta { S }_{ t }=log\left( { S }_{ t }/{ S }_{ t-1 } \right)$ to represent the percent change in species diversity of the catch between years and $\Delta { D }_{ t }=log\left( { D }_{ t }/{ D }_{ t-1 } \right)$ to represent the percent change in fishing effort (days fished) between years. The model for the the expected log revenue of individual *i* in year *t*, $$\widehat { log\left( { R }_{ i,t } \right)  } ={ B }_{ 0,j,t }+{ B }_{ 1,j,t }\cdot \Delta S_{ t }+{ B }_{ 2,j,t }\cdot \left( \Delta S_{ t }\cdot { S }_{ t-1 } \right) +{ B }_{ 3,j }\cdot \Delta D_{ t }+log\left( { R }_{ i,t-1 } \right) $$, where *j* subscripts the salmon permit (Table 1). We model the time varying permit specific intercepts as random and exchangeable across years, ${ B }_{ 0,j,t }\sim Normal\left( { u }_{ { B }_{ 0 } },{ \sigma  }_{ { B }_{ 0 } } \right)$. The term ${ B }_{ 1,j,t }$ represents the permit level random effect of diversifying species. Our model extendeds that of [@anderson_benefits] to include the interaction terms $\Delta S_{ t }\cdot { S }_{ t-1 }$ which allow the effect of diversification to be dependent on the previous level of diversification. We assume this relationship is linear, and the coefficients ${ B }_{ 1,j,t }$ and ${ B }_{ 2,j,t }$ allow this to vary by permit and through time. The flexibility in these terms allows the benefit of increasing species diversity to be positive in some years (generalizing benefits revenue), negative in other years (specializing benefits revenue) and neutral in other years (revenue is independent alterations to catch composition). We constrain the random effects associated with diversification to be modeled as autoregressive processes with shared variances across strategies, for example: ${ B }_{ 1,j,t }\sim Normal\left( { B }_{ 1,j,t-1 },{ \sigma  }_{ { B }_{ 1 } } \right)$. Like [@anderson_benefits] we also allow changing effort (measured as percent change in days fished) ${ B }_{ 4,j }$ to vary by strategy, but we do not allow this effect to vary through time. Finally, the term $log\left( { R }_{ i,t-1 } \right)$ is included as an offset [@anderson_benefits].

- Fit models from Sean's paper, but include time varying interactions on species diversity.
All done in STAN [@stan_development_team_stan_2015] using the NUTS algorithm [@hoffman_no-u-turn_2014].

@@

Figure 1 shows that the time varying intercepts on the mean, also in [@anderson_benefits] are basically flat. These represent common fluctuations in revenue experienced by all permit holders.

Figure 2 is largley illustrative, but shows the benefit of including time varying coefficients.

Figure 3 shows the temporal trends, by permit, in going from 1 to 2 species fished, and how those effects have changed through time. S01A, S01E, S03A appear to be decreasing. S01K and S01M increasing. S03E pretty flat?

Figure 4 shows that temporal trends, by permit, in going from 1 to 2 species fished on the variance. From [@anderson_benefits] we'd expect diversification to redudce variance -- that seems to be the biggest signal for S01K and S01M in the years of low sockeye prices.

@@

- Prices are generally synchronous for all spp (especially coho/chum/pink), and have increased at a faster rate compared to chum and pink salmons since 2000 (Fig A8-A9). We're also seeing a trend toward greater synchrony between species (Fig A11-A12, A14). Projecting future prices are complex [@knapp_trends]. Synchrony in prices may largely be function of marketing, improved product, freezing, etc.

@@

```{r, echo=FALSE, fig.pos="placeHere", fig.cap="Time varying intercepts on estimated mean revenue, by permit."}
library(rstan)

# grab original data to look at permit names
diffdat = readRDS(file="../../salmon/data-generated/salmon.rds")
years = seq(min(diffdat$year), max(diffdat$year))
permits = levels(as.factor(diffdat$permit))

yearpermit = data.frame("yrpermit"=levels(diffdat$year_permit))
yearpermit$year = as.numeric(substr(yearpermit$yrpermit,1,4))
yearpermit$permit = substr(yearpermit$yrpermit,6,9)

load("../../salmon/analysis/model-linear.Rdata")

est = extract(mod, permute="TRUE")

yearpermit$b0 = apply(est$b0_str_yr,2,median)
yearpermit$b0_low = apply(est$b0_str_yr,2,quantile,0.25)
yearpermit$b0_hi = apply(est$b0_str_yr,2,quantile,0.75)

ggplot(data=yearpermit) + 
  geom_ribbon(aes(x=year,ymin=b0_low,ymax=b0_hi),alpha=0.3) +
  geom_line(aes(x=year,y=b0)) + 
  facet_wrap(~permit) + xlab("Year") + ylab("Intercepts on mean")
